---
title: "Gov 50 Final Project"
author: "Amber Nolazco-Torres"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


```{r}
library(tidyverse)
schools <- read_csv("crdc-2015-16-ca-aug-2018.csv")
schools
```

```{r}
subset <- schools  %>% 
  filter(SCH_GRADE_G09 == "YES" & SCH_GRADE_G10 == "YES" & SCH_GRADE_G11 == "YES" & SCH_GRADE_G12 == "YES" &    SCH_GRADE_G08 == "NO") 

subset[subset < 0] <- NA 
  
median(subset$SCH_APCOURSES, na.rm = TRUE) 




```

## Final Project Milestone 4

```{r}
## schools[schools == -9] <- NA

subset <- subset |> 
  filter(!is.na(SCH_APEXAM_ONEORMORE_HP_M) & !is.na(SCH_APEXAM_ONEORMORE_HI_F)) |> 
  mutate(total_hispanic_students_AP = (SCH_APEXAM_ONEORMORE_HP_M + SCH_APEXAM_ONEORMORE_HI_F),
         total_hispanic_student_enrollment = (SCH_ENR_HI_M + SCH_ENR_HI_F)) |> 
  mutate(school_level_hisp_m_AP_rate = total_hispanic_students_AP/total_hispanic_student_enrollment)
 

subset_hispanic <- subset |>
  mutate(
    AP_distribution = case_when(
      SCH_APCOURSES >= 12 ~ "Schools with 12 or more AP Classes", 
      SCH_APCOURSES < 12 & SCH_APCOURSES != 0 ~ "Schools with less than 12 AP Classes", 
      is.na(SCH_APCOURSES) ~ "No AP Classes")) |>
  group_by(AP_distribution) %>% 
  summarize(Proportion_APClasses = mean(school_level_hisp_m_AP_rate, na.rm = TRUE)) |>
  ggplot(mapping = aes(x = AP_distribution, y = Proportion_APClasses )) +
  geom_col() +
  labs(x = "Number of AP Classes Offered (CA Public High Schools)",
       y = "Proportion of Hispanic Students in at least 1 AP Class",
       title = "California: Proportion of Hispanic Students in at least 1 AP Class")
subset_hispanic

subset_groups <- subset |>
  mutate(
    AP_distribution = case_when(
      SCH_APCOURSES >= 12 ~ "Schools with 12 or more AP Classes", 
      SCH_APCOURSES < 12 & SCH_APCOURSES != 0 ~ "Schools with less than 12 AP Classes", 
      is.na(SCH_APCOURSES) ~ "No AP Classes")) |>
  group_by(AP_distribution) %>% 
  summarize(Proportion_APClasses = mean(school_level_hisp_m_AP_rate, na.rm = TRUE)) |> 
  pivot_wider(names_from = AP_distribution, values_from = Proportion_APClasses) |> 
  mutate(ATE = `Schools with 12 or more AP Classes` - `Schools with less than 12 AP Classes`)

subset_groups

```

In the graph above I compared the proportion of Hispanic students taking at least 1 AP class in California public schools. In order to better account for differences in school funding I created one group for Hispanic students who attend a school that offers "12 or more AP classes" and another for students who attend a school that offers "less than 12 AP classes". Twelve was the average number of AP classes offered by the high schools in the data set. 8.997% of Hispanic students in California public schools that offer 12 or more AP classes are taking at least 1 AP class. 8.081% of Hispanic students in California public schools that offer less than 12 AP classes are taking at least 1 AP class. The ATE between these two groups is 0.0916%. Hispanic students in schools that offer more than 12 AP classes are slightly more likely to be enrolled in at least 1 AP class. It is important to mention that there are possible confounding variables. The funding of schools and the quality of AP classes may impact the percentage of Hispanic students enrolled in the courses.




